---
title: "MenstruLife prevalence paper (Aim 1)"
author: "Sam"
date: "2026-05-10"
output: html_document
---
# Load datasets
```{r setup, include=FALSE}
library(haven)
bcs_der <- read_dta("C:/Users/sk1014/OneDrive - University of Exeter/Sharp, Gemma's files - Shared/Analysis plans/BCS1970/Merged sweeps/bcs_der.dta")

abcd_der <- read_dta("C:/Users/sk1014/OneDrive - University of Exeter/Sharp, Gemma's files - Shared/Analysis plans/ABCD/abcd_der_wide.dta")

raine_der <- read_dta("C:/Users/sk1014/OneDrive - University of Exeter/Sharp, Gemma's files - Shared/Analysis plans/Raine/raine_der.dta")

alspac_g0_der <- read_dta("C:/Users/sk1014/OneDrive - University of Exeter/Sharp, Gemma's files - Shared/Analysis plans/ALSPAC/alspac_g0_der_point.dta")

alspac_g1_der <- read_dta("C:/Users/sk1014/OneDrive - University of Exeter/Sharp, Gemma's files - Shared/Analysis plans/ALSPAC/alspac_g1_der.dta")
```

#Function to check HMB/MP available at which ages
```{r}
get_age_availability <- function(data, cohort_name){

  df <- data %>%
    select(starts_with("hmb_age"), starts_with("mp_age")) %>%
    pivot_longer(
      cols = everything(),
      names_to = "var",
      values_to = "value"
    ) %>%
    mutate(
      symptom = ifelse(grepl("^hmb_age", var), "HMB", "MP"),
      age = as.numeric(gsub("[^0-9]", "", var))
    ) %>%
    group_by(symptom, age) %>%
    summarise(
      available = ifelse(all(is.na(value)), 0, 1),
      .groups = "drop"
    ) %>%
    mutate(cohort = cohort_name)

  return(df)
}

```

#Sample plot (Figure 1)
```{r}
library(tidyverse)
library(stringr)

moba_heat <- tibble(
  cohort = "MoBa",
  age = 20,
  symptom = c("HMB", "MP"),
  available = 1
)

heat_df <- bind_rows(
  get_age_availability(abcd_der, "ABCD"),
  get_age_availability(alspac_g1_der, "ALSPAC G1"),
  get_age_availability(raine_der, "RAINE"),
  moba_heat,
  get_age_availability(bcs_der, "BCS70"),
  get_age_availability(alspac_g0_der, "ALSPAC G0")
)

heat_df <- heat_df %>%
  mutate(
    cohort = recode(
      cohort,
      "ABCD" = "ABCD\n(N=5,112)",
      "ALSPAC G1" = "ALSPAC G1\n(N=4,247)",
      "RAINE" = "RAINE\n(N=842)",
      "MoBa" = "MoBa\n(N=5,342)",
      "BCS70" = "BCS70\n(N=3,687)",
      "ALSPAC G0" = "ALSPAC G0\n(N=11,072)"
    )
  )

heat_df$cohort <- factor(
  heat_df$cohort,
  levels = c(
    "ALSPAC G0\n(N=11,072)",
    "BCS70\n(N=3,687)",
    "MoBa\n(N=5,342)",
    "RAINE\n(N=842)",
    "ALSPAC G1\n(N=4,247)",
    "ABCD\n(N=5,112)"
  )
)

heat_df <- heat_df %>%
  mutate(
    stage = case_when(
      age < 20 ~ "Adolescence",
      age >= 20 & age < 45 ~ "Mid reproductive life",
      age >= 45 ~ "Late reproductive life"
    )
  )

heat_df$stage <- factor(
  heat_df$stage,
  levels = c(
    "Adolescence",
    "Mid reproductive life",
    "Late reproductive life"
  )
)

heat_df <- heat_df %>%
  group_by(cohort, symptom, stage) %>%
  group_modify(~{
    
    ages <- switch(
      as.character(.y$stage),
      "Adolescence" = 10:19,
      "Mid reproductive life" = 20:44,
      "Late reproductive life" = 45:56
    )
    
    tidyr::complete(
      .x,
      age = ages,
      fill = list(available = 0)
    )
    
  }) %>%
  ungroup()

ggplot(heat_df) +

  geom_tile(
    aes(
      x = age,
      y = cohort,
      fill = factor(available)
    ),
    width = 0.9,
    height = 0.75,
    colour = "white",
    linewidth = 0.25
  ) +

  facet_grid(
    rows = vars(symptom),
    cols = vars(stage),
    scales = "free_x",
    space = "free_x"
  ) +

  scale_fill_manual(
    values = c(
      "0" = "#F0F0F0",
      "1" = "#D55E00"
    ),
    guide = "none"
  ) +

scale_x_continuous(
  breaks = scales::breaks_pretty(5),
  expand = c(0, 0)
) +

  labs(
    x = "Age (years)",
    y = "Cohort"
  ) +

  theme_minimal(base_size = 13) +

  theme(
    panel.grid = element_blank(),

    strip.text = element_text(
      size = 11,
      face = "bold"
    ),

    axis.title = element_text(
      face = "bold"
    ),

    axis.text.y = element_text(
      size = 11
    ),

    axis.text.x = element_text(
      size = 10
    ),

    panel.spacing.x = unit(1.5, "lines"),
    panel.spacing.y = unit(1.2, "lines")
  )


ggsave("sample_agerange_by_cohorts.png", width = 12, height = 7, dpi = 300)

```

# Cumulative Prevalence
#Functions for cumulative prevalence
#HMB only, MP only, and Co-occurance
```{r, echo=TRUE, results='asis', fig.show='hold'}

calculate_lifetime_prevalence <- function(data, id_var = "participant_id", var_prefix = "hmb_mp") {
  
  data <- data %>% dplyr::mutate(across(starts_with(var_prefix), haven::zap_labels))
  
  long_data <- data %>%
    dplyr::select(all_of(id_var), dplyr::starts_with(var_prefix)) %>%
    tidyr::pivot_longer(
      cols = dplyr::starts_with(var_prefix),
      names_to = "age_var",
      values_to = "hmb_mp"
    ) %>%
    dplyr::mutate(age = as.integer(gsub("[^0-9]", "", age_var)))
  
  lifetime_status <- long_data %>%
    dplyr::group_by(!!rlang::sym(id_var)) %>%
    dplyr::summarise(max_status = max(hmb_mp, na.rm = TRUE), .groups = "drop") %>%
    dplyr::mutate(
      status = dplyr::case_when(
        max_status == 0 ~ "Neither",
        max_status == 1 ~ "HMB only",
        max_status == 2 ~ "MP only",
        max_status == 3 ~ "HMB & MP",
        TRUE ~ NA_character_
      )
    ) %>%
    dplyr::filter(!is.na(status))
  
lifetime_summary <- lifetime_status %>%
  group_by(status) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(
    total = sum(count),
    prevalence = count / total
  )

ci <- binom::binom.wilson(
  lifetime_summary$count,
  lifetime_summary$total
)

lifetime_summary <- bind_cols(
  lifetime_summary,
  ci %>% select(lower, upper)
) %>%
  mutate(
    percent = round(prevalence * 100, 1),
    lower_ci = round(lower * 100, 1),
    upper_ci = round(upper * 100, 1)
  )
  
  # Plot
  p <- ggplot(lifetime_summary, aes(x = reorder(status, -percent), y = percent, fill = status)) +
    geom_col(color = "grey40", width = 0.6) +
    geom_text(aes(label = paste0(percent, "%")), vjust = -0.4, size = 4) +
    scale_fill_brewer(palette = "Set2") +
    labs(
      title = "Cumulative Prevalence of HMB and/or MP",
      x = NULL,
      y = "Percentage (%)"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      legend.position = "none",
      panel.grid.major.x = element_blank(),
      plot.title = element_text(face = "bold", size = 16)
    )
  
  return(list(summary = lifetime_summary, plot = p))
}


```

#HMB/MP Ever
```{r}
calculate_symptom_ever <- function(data, id_var = "participant_id") {

  library(dplyr)

  hmb_vars <- grep("^hmb_age", names(data), value = TRUE)
  mp_vars  <- grep("^mp_age", names(data), value = TRUE)

  data_ever <- data %>%
    mutate(
      hmb_ever = ifelse(rowSums(select(., all_of(hmb_vars)) == 1, na.rm = TRUE) > 0, 1, 0),
      mp_ever  = ifelse(rowSums(select(., all_of(mp_vars)) == 1, na.rm = TRUE) > 0, 1, 0),
      both_ever = ifelse(hmb_ever == 1 | mp_ever == 1, 1, 0)
    )

summary_table <- tibble(
  symptom = c("HMB ever","MP ever","Either ever"),
  count = c(
    sum(data_ever$hmb_ever, na.rm=TRUE),
    sum(data_ever$mp_ever, na.rm=TRUE),
    sum(data_ever$both_ever, na.rm=TRUE)
  ),
  total = nrow(data_ever)
) %>%
  mutate(
    prevalence = count/total
  )

ci <- binom::binom.wilson(
  summary_table$count,
  summary_table$total
)

summary_table <- bind_cols(
  summary_table,
  ci %>% select(lower, upper)
) %>%
  mutate(
    percent = round(prevalence*100,1),
    lower_ci = round(lower*100,1),
    upper_ci = round(upper*100,1)
  )

  return(list(
    data = data_ever,
    summary = summary_table
  ))
}
```

## Cumulative prevalence calculation

```{r, echo=TRUE, results='asis', fig.show='hold'}
library(dplyr)
library(tidyr)
library(ggplot2)
library(haven)
library(rlang)

bcs_lifetime_prev <- calculate_lifetime_prevalence(data = bcs_der, id_var = "bcsid", var_prefix = "hmb_mp")

abcd_lifetime_prev <- calculate_lifetime_prevalence(data = abcd_der, id_var = "participant_id", var_prefix = "hmb_mp")

raine_lifetime_prev <- calculate_lifetime_prevalence(data = raine_der, id_var = "ID", var_prefix = "hmb_mp")

alspac_g0_lifetime_prev <- calculate_lifetime_prevalence(data = alspac_g0_der, id_var = "cidB4396", var_prefix = "hmb_mp")

alspac_g1_lifetime_prev <- calculate_lifetime_prevalence(data = alspac_g1_der, id_var = "cidB4396", var_prefix = "hmb_mp")

```

#Combined cumulative plots
#HMB only, MP only and Both (Not included in paper)
```{r}

combined_lifetime <- bind_rows(
  bcs_lifetime_prev$summary %>% mutate(cohort = "BCS70"),
  abcd_lifetime_prev$summary %>% mutate(cohort = "ABCD"),
  raine_lifetime_prev$summary %>% mutate(cohort = "RAINE"),
  alspac_g0_lifetime_prev$summary %>% mutate(cohort = "ALSPAC G0"),
  alspac_g1_lifetime_prev$summary %>% mutate(cohort = "ALSPAC G1")
)

combined_lifetime$cohort <- factor(
  combined_lifetime$cohort,
  levels = c("ABCD", "ALSPAC G1", "RAINE", "BCS70", "ALSPAC G0")
)

ggplot(combined_lifetime, aes(x = cohort, y = percent, fill = status)) +
  geom_col(position = position_dodge(width = 0.7), 
           color = "grey40", width = 0.6) +
  geom_errorbar(
  aes(
    ymin = lower_ci,
    ymax = upper_ci
  ),
  position = position_dodge(width = 0.7),
  width = 0.2,
  linewidth = 0.4
) +
  geom_text(
  aes(
    y = upper_ci + 1.4,
    label = paste0(percent, "%")
  ),
  position = position_dodge(width = 0.7),
  size = 3.8
) +
  scale_fill_manual(values = c(
  "Neither"   = "#E0E0E0",
  "MP only"   = "#A6A6A6",
  "HMB only"  = "#737373",
  "HMB & MP"  = "grey25"
)) +
  labs(
    x = "Cohort",
    y = "Percentage (%)",
    fill = "Symptom Status"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 16)
  )
ggsave("cumulative_prevalence_by_cohorts.png", width = 12, height = 6, dpi = 300)

```

#HMB/MP plot (Figure 2)
```{r}
bcs_symptom_ever <- calculate_symptom_ever(data = bcs_der, id_var = "bcsid")
bcs_symptom_ever$summary

abcd_symptom_ever <- calculate_symptom_ever(abcd_der, id_var = "participant_id")
abcd_symptom_ever$summary

raine_symptom_ever <- calculate_symptom_ever(raine_der, id_var = "ID")
raine_symptom_ever$summary

alspac_g0_symptom_ever <- calculate_symptom_ever(alspac_g0_der, id_var = "cidB4396")
alspac_g0_symptom_ever$summary

alspac_g1_symptom_ever <- calculate_symptom_ever(alspac_g1_der, id_var = "cidB4396")
alspac_g1_symptom_ever$summary

ever_combined <- bind_rows(
  bcs_symptom_ever$summary %>% mutate(cohort = "BCS70"),
  abcd_symptom_ever$summary %>% mutate(cohort = "ABCD"),
  raine_symptom_ever$summary %>% mutate(cohort = "RAINE"),
  alspac_g0_symptom_ever$summary %>% mutate(cohort = "ALSPAC G0"),
  alspac_g1_symptom_ever$summary %>% mutate(cohort = "ALSPAC G1")
)

ever_combined

unique(ever_combined$symptom)
ever_combined <- ever_combined %>%
  mutate(
    symptom = trimws(symptom)   
  )

ever_combined$cohort <- factor(
  ever_combined$cohort,
  levels = c("ABCD","ALSPAC G1","RAINE","BCS70","ALSPAC G0")
)
ever_combined$symptom <- factor(
  ever_combined$symptom,
  levels = c(
    "HMB ever",
    "MP ever",
    "Either ever"
  )
)

ever_plot <- ggplot(
  ever_combined,
  aes(x = cohort, y = percent, fill = symptom)
) +

  geom_col(
    position = position_dodge(width = 0.8),
    width = 0.7,
    color = "grey30"
  ) +
geom_errorbar(
  aes(
    ymin = lower_ci,
    ymax = upper_ci
  ),
  position = position_dodge(width = 0.8),
  width = 0.15,
  linewidth = 0.4
) +
geom_text(
  aes(
    y = upper_ci + 1.8,
    label = paste0(percent, "%")
  ),
  position = position_dodge(width = 0.8),
  size = 3.8
) +

  scale_fill_manual(
    values = c(
      "HMB ever" = "grey30",
      "MP ever" = "grey55",
      "Either ever" = "grey75"
    )
  ) +
scale_y_continuous(
  limits = c(0,100),
  expand = expansion(mult = c(0,0.02))
) +
  labs(
    x = "Cohort",
    y = "Percentage (%)",
    fill = "Symptom status"
  ) +

  theme_minimal(base_size = 14) +

  theme(
    legend.position = "bottom",
    panel.grid.major.x = element_blank(),
    plot.title = element_text(face = "bold", size = 16)
  )

ever_plot

ggsave("symptom_ever_by_cohorts.png", width = 10, height = 6, dpi = 300)


```

#Cumulative prevalence summary table
```{r}
ever_summary <- bind_rows(
  abcd_symptom_ever$summary %>% mutate(Cohort = "ABCD"),
  alspac_g1_symptom_ever$summary %>% mutate(Cohort = "ALSPAC G1"),
  raine_symptom_ever$summary %>% mutate(Cohort = "RAINE"),
  bcs_symptom_ever$summary %>% mutate(Cohort = "BCS70"),
  alspac_g0_symptom_ever$summary %>% mutate(Cohort = "ALSPAC G0")
) %>%
  select(Cohort, everything())
lifetime_summary <- bind_rows(
  abcd_lifetime_prev$summary %>% mutate(Cohort = "ABCD"),
  alspac_g1_lifetime_prev$summary %>% mutate(Cohort = "ALSPAC G1"),
  raine_lifetime_prev$summary %>% mutate(Cohort = "RAINE"),
  bcs_lifetime_prev$summary %>% mutate(Cohort = "BCS70"),
  alspac_g0_lifetime_prev$summary %>% mutate(Cohort = "ALSPAC G0")
) %>%
  select(Cohort, everything())
library(writexl)

write_xlsx(list("Ever prevalence" = ever_summary,"Lifetime status" = lifetime_summary),"Cumulative_prevalence_with_CIs.xlsx")
```


#Point prevalence 
#Function for Point Prevalence

```{r, echo=TRUE, results='asis', fig.show='hold'}

calculate_point_prevalence <- function(data, id_var = "bcsid", var_prefix = "hmb_mp") {
  
  hmb_mp_long <- data %>%
    dplyr::select(all_of(id_var), starts_with(var_prefix)) %>%
    pivot_longer(
      cols = starts_with(var_prefix),
      names_to = "age",
      values_to = "hmb_mp"
    ) %>%
    mutate(
      hmb_mp = suppressWarnings(as.numeric(hmb_mp)),  
      age = as.integer(gsub("[^0-9]", "", age))
    ) %>%
    filter(!is.na(age)) %>%
    filter(hmb_mp %in% 0:3) %>%  
    mutate(
      status = dplyr::case_when(
        hmb_mp == 0 ~ "Neither",
        hmb_mp == 1 ~ "HMB only",
        hmb_mp == 2 ~ "MP only",
        hmb_mp == 3 ~ "HMB & MP"
      )
    )
  
  point_prev <- hmb_mp_long %>%
    group_by(age, status) %>%
    summarise(n = n(), .groups = "drop") %>%
    group_by(age) %>%
    mutate(
      total = sum(n),
      Prevalence = n / total,
      SE = sqrt((Prevalence * (1 - Prevalence)) / total),
      lci = Prevalence - 1.96 * SE,
      uci = Prevalence + 1.96 * SE
    ) %>%
    ungroup() %>%
    dplyr::select(age, status, n, Prevalence, SE, lci, uci)
  
  cat("\nPoint Prevalence of HMB and/or MP by age:\n\n")
  print(point_prev)
  
  #Plots
  p1 <- ggplot(point_prev, aes(x = age, y = Prevalence, ymin = lci, ymax = uci)) +
    geom_errorbar(width = 0.5) +
    geom_point(size = 1) +
    facet_wrap(~ status) +
    labs(
      title = "Prevalence of HMB/MP by Age with 95% CI",
      x = "Age",
      y = "Prevalence"
    ) +
    theme_minimal()
  
  p2 <- ggplot(point_prev, aes(x = factor(age), y = Prevalence, fill = status)) +
    geom_col(position = position_dodge(width = 0.9)) +
    geom_text(
      aes(label = paste0(round(Prevalence * 100, 1), "%")),
      position = position_dodge(width = 0.9),
      vjust = -0.3, size = 3
    ) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    labs(
      title = "Point prevalence by age",
      x = "Age",
      y = "Prevalence (%)",
      fill = "Status"
    ) +
    theme_minimal()
  
  p3 <- ggplot(point_prev, aes(x = age, y = Prevalence, group = status, color = status)) +
    geom_point() +
    geom_smooth(method = "loess", se = FALSE) +
    labs(
      title = "Point Prevalence by Age",
      x = "Age",
      y = "Prevalence",
      color = "Status"
    ) +
    theme_minimal()
  
  p4 <- ggplot(point_prev, aes(x = age, y = Prevalence, ymin = lci, ymax = uci, color = status)) +
    geom_errorbar(width = 0.4) +
    geom_point(size = 1) +
    labs(
      title = "Prevalence with 95% CI by Age and Status",
      x = "Age",
      y = "Prevalence",
      color = "Status"
    ) +
    theme_minimal()
  
  # Return results
  invisible(list(
    prevalence_table = point_prev,
    plot_facet = p1,
    plot_bar = p2,
    plot_trend = p3,
    plot_errorbar = p4
  ))
}

  
```

#Point Prevalence calculation
```{r, echo=TRUE, results='asis', fig.show='hold'}

bcs_point_prev <- calculate_point_prevalence(data = bcs_der, id_var = "bcsid", var_prefix = "hmb_mp")

abcd_point_prev<- calculate_point_prevalence(data = abcd_der, id_var = "participant_id", var_prefix = "hmb_mp")

raine_point_prev<- calculate_point_prevalence(data = raine_der, id_var = "ID", var_prefix = "hmb_mp")

alspac_g0_point_prev<- calculate_point_prevalence(data = alspac_g0_der, id_var = "cidB4396", var_prefix = "hmb_mp")

alspac_g1_point_prev<- calculate_point_prevalence(data = alspac_g1_der, id_var = "cidB4396", var_prefix = "hmb_mp")

```


# Combine prevalence tables from each cohort
#HMB only,MP only, and Both
```{r}
library(dplyr)
library(ggplot2)

moba_point <- tibble(cohort = "MoBa",age = 20, status = c("HMB & MP", "HMB only", "MP only"), 
                     percent = c(21.7, 3.8, 52.2)) #Based on the results sent from MoBa team

combined_point <- bind_rows(
  bcs_point_prev$prevalence_table %>% mutate(cohort = "BCS70"),
  abcd_point_prev$prevalence_table %>% mutate(cohort = "ABCD"),
  raine_point_prev$prevalence_table %>% mutate(cohort = "RAINE"),
  alspac_g0_point_prev$prevalence_table %>% mutate(cohort = "ALSPAC G0"),
  alspac_g1_point_prev$prevalence_table %>% mutate(cohort = "ALSPAC G1")
)
combined_point <- combined_point %>%
  filter(status %in% c("HMB only", "MP only", "HMB & MP")) %>%
  mutate(percent = Prevalence * 100)

combined_point <- bind_rows(  combined_point, moba_point)

combined_point$age <- as.numeric(combined_point$age)

combined_point$status <- factor(
  combined_point$status,
  levels = c("HMB only", "MP only", "HMB & MP")
)

combined_point$cohort <- factor(
  combined_point$cohort,
  levels = c("ABCD", "ALSPAC G1", "MoBa", "RAINE", "BCS70", "ALSPAC G0")
)

```

#HMB/MP point prev 
```{r}
calculate_point_any_correct <- function(data, cohort_name){

  library(dplyr)
  library(tibble)

  hmb_vars <- grep("^hmb_age", names(data), value = TRUE)
  mp_vars  <- grep("^mp_age", names(data), value = TRUE)

  ages <- unique(gsub("[^0-9]", "", hmb_vars))

  results <- lapply(ages, function(a){

    hmb_var <- paste0("hmb_age", a)
    mp_var  <- paste0("mp_age", a)

    if(!(hmb_var %in% names(data) & mp_var %in% names(data))) return(NULL)

    df <- data

    # HMB
    hmb_obs <- !is.na(df[[hmb_var]])
    n_hmb_total <- sum(hmb_obs)
    n_hmb <- sum(df[[hmb_var]] == 1, na.rm = TRUE)
    hmb_prev <- n_hmb / n_hmb_total

    # MP
    mp_obs <- !is.na(df[[mp_var]])
    n_mp_total <- sum(mp_obs)
    n_mp <- sum(df[[mp_var]] == 1, na.rm = TRUE)
    mp_prev <- n_mp / n_mp_total

    # ANY
    any_obs <- hmb_obs | mp_obs
    n_any_total <- sum(any_obs)
    n_any <- sum(
      (df[[hmb_var]] == 1 | df[[mp_var]] == 1),
      na.rm = TRUE
    )
    any_prev <- n_any / n_any_total

    tibble(
      cohort = cohort_name,
      age = as.numeric(a),

      # counts
      n_HMB = n_hmb,
      n_MP  = n_mp,
      n_any = n_any,

      # denominators (optional but useful)
      denom_HMB = n_hmb_total,
      denom_MP  = n_mp_total,
      denom_any = n_any_total,

      # percentages
      HMB = hmb_prev * 100,
      MP  = mp_prev * 100,
      `HMB or MP` = any_prev * 100
    )
  })

  bind_rows(results)
}
abcd_any   <- calculate_point_any_correct(abcd_der, "ABCD")
g1_any     <- calculate_point_any_correct(alspac_g1_der, "ALSPAC G1")
raine_any  <- calculate_point_any_correct(raine_der, "RAINE")
bcs_any    <- calculate_point_any_correct(bcs_der, "BCS70")
g0_any     <- calculate_point_any_correct(alspac_g0_der, "ALSPAC G0")

moba_any <- tibble(
  cohort = "MoBa",
  age = 20,
  HMB = 25.1,
  MP = 73.9,
  `HMB or MP` = 77.2
)


combined_any <- bind_rows(abcd_any, g1_any, raine_any, moba_any, bcs_any, g0_any)

combined_any_long <- combined_any %>%
  pivot_longer(
    cols = c(HMB, MP, `HMB or MP`),
    names_to = "status",
    values_to = "percent"
  )

combined_any_long <- combined_any_long %>%
  mutate(
    age = as.numeric(age),
    status = factor(status, levels = c("HMB","MP","HMB or MP")),
    cohort = factor(cohort, levels = c("ABCD","ALSPAC G1","RAINE","MoBa", "BCS70","ALSPAC G0"))
  )

```


#Tables- cumulative prev, point prev- hmb only, mp only, hmb&mp, hmb, mp, hmb/mp
```{r}
lifetime_table <- bind_rows(
  bcs_lifetime_prev$summary %>% mutate(cohort = "BCS70"),
  abcd_lifetime_prev$summary %>% mutate(cohort = "ABCD"),
  raine_lifetime_prev$summary %>% mutate(cohort = "RAINE"),
  alspac_g0_lifetime_prev$summary %>% mutate(cohort = "ALSPAC G0"),
  alspac_g1_lifetime_prev$summary %>% mutate(cohort = "ALSPAC G1")
) %>%

  filter(status != "Neither") %>%

  mutate(
    percent = round(percent, 1),
    value = paste0(count, " (", percent, "%)")
  ) %>%

  select(cohort, status, value) %>%

  pivot_wider(
    names_from = status,
    values_from = value
  ) %>%

  mutate(
    cohort = factor(cohort,
      levels = c("ABCD","ALSPAC G1","RAINE","BCS70","ALSPAC G0")
    )
  ) %>%

  arrange(cohort)

lifetime_table

combined_point <- bind_rows(
  bcs_point_prev$prevalence_table %>% mutate(cohort = "BCS70"),
  abcd_point_prev$prevalence_table %>% mutate(cohort = "ABCD"),
  raine_point_prev$prevalence_table %>% mutate(cohort = "RAINE"),
  alspac_g0_point_prev$prevalence_table %>% mutate(cohort = "ALSPAC G0"),
  alspac_g1_point_prev$prevalence_table %>% mutate(cohort = "ALSPAC G1")
)
point_table <- combined_point %>%

  filter(status %in% c("HMB only","MP only","HMB & MP", "Neither")) %>%

  mutate(
    percent = round(Prevalence * 100, 1),
    value = paste0(n, " (", percent, ")")
  ) %>%

  select(cohort, age, status, value) %>%

  pivot_wider(
    names_from = status,
    values_from = value
  ) %>%

  mutate(
    cohort = factor(cohort,
      levels = c("ABCD","ALSPAC G1","RAINE","BCS70","ALSPAC G0")
    )
  ) %>%

  arrange(cohort, age)

point_table

point_prev_any_formatted <- combined_any %>%

  mutate(
    HMB = paste0(n_HMB, " (", round(HMB,1), ")"),
    MP  = paste0(n_MP,  " (", round(MP,1), ")"),
    `HMB or MP` = paste0(n_any, " (", round(`HMB or MP`,1), ")")
  ) %>%

  select(cohort, age, HMB, MP, `HMB or MP`) %>%

  mutate(
    cohort = factor(cohort,
      levels = c("ABCD","ALSPAC G1","RAINE","BCS70","ALSPAC G0")
    )
  ) %>%

  arrange(cohort, age)

point_prev_any_formatted

write.csv(lifetime_table, "lifetime_prevalence_table.csv", row.names = FALSE)

write.csv(point_table, "point_prevalence_table.csv", row.names = FALSE)

write.csv(point_prev_any_formatted,"point_prevalence_hmb_mp_any_by_age.csv",row.names = FALSE)


```

#For supplementary table
```{r}

# HMB only / MP only / Both
status_counts <- combined_point %>%

  filter(status %in% c("HMB only", "MP only", "HMB & MP")) %>%

  select(
    cohort,
    age,
    status,
    n
  ) %>%

  pivot_wider(
    names_from = status,
    values_from = n
  ) %>%

  rename(
    Both = `HMB & MP`
  )

# HMB any / MP any / Either / Total / Neither
any_counts <- combined_any %>%

  transmute(
    cohort,
    age,
    HMB_any = n_HMB,
    MP_any = n_MP,
    Either = n_any,
    Total = denom_any,
    Neither = denom_any - n_any
  )

# Merge
point_prevalence_table <- left_join(
  any_counts,
  status_counts,
  by = c("cohort", "age")
)

# Add MoBa
moba_row <- tibble(
  cohort = "MOBA",
  age = 20,
  HMB_any = 1340,
  MP_any = 3946,
  Either = 4126,
  Both = 1160,
  `HMB only` = 180,
  `MP only` = 2786,
  Total = 5342,
  Neither = 1216
)

point_prevalence_table <- bind_rows(
  point_prevalence_table,
  moba_row
)

point_prevalence_table <- point_prevalence_table %>%
  mutate(
    Both = tidyr::replace_na(Both, 0),
    `HMB only` = tidyr::replace_na(`HMB only`, 0),
    `MP only` = tidyr::replace_na(`MP only`, 0)
  )

point_prevalence_table <- point_prevalence_table %>%

  mutate(

    `Age group` = case_when(
      age < 20 ~ "Ado",
      age >= 20 & age < 45 ~ "Mid",
      age >= 45 ~ "Peri"
    ),

    cohort = factor(
      cohort,
      levels = c(
        "ABCD",
        "ALSPAC G1",
        "RAINE",
        "MOBA",
        "BCS70",
        "ALSPAC G0"
      )
    )

  ) %>%

  arrange(cohort, age)

# Final table
point_prevalence_table <- point_prevalence_table %>%

  select(
    Cohort = cohort,
    `Age group`,
    Age = age,
    `HMB any` = HMB_any,
    `MP any` = MP_any,
    Either,
    Both,
    `HMB only`,
    `MP only`,
    Total,
    Neither
  )

# Export
write.csv(point_prevalence_table,"point_prevalence_counts.csv", row.names = FALSE)
```


#Combined plot for point prev (Figure 3)
```{r}

dat <- read.csv("C:/Users/sk1014/OneDrive - University of Exeter/Sharp, Gemma's files - Shared/Paper/Aim 1 paper/point_prevalence_counts.csv")


library(dplyr)
library(tidyr)
library(ggplot2)
install.packages("binom")
library(binom)
install.packages("ggh4x")
library(ggh4x)
library(patchwork)

prev_dat <- bind_rows(
  
  # HMB
  dat %>%
    transmute(
      Cohort,
      Age,
      Age.group,
      Total,
      Symptom = "HMB",
      Type = "Any",
      Cases = HMB.any
    ),
  
  dat %>%
    transmute(
      Cohort,
      Age,
      Age.group,
      Total,
      Symptom = "HMB",
      Type = "Isolated",
      Cases = HMB.only
    ),
  
  # MP
  dat %>%
    transmute(
      Cohort,
      Age,
      Age.group,
      Total,
      Symptom = "MP",
      Type = "Any",
      Cases = MP.any
    ),
  
  dat %>%
    transmute(
      Cohort,
      Age,
      Age.group,
      Total,
      Symptom = "MP",
      Type = "Isolated",
      Cases = MP.only
    ),
  
  # BOTH
  dat %>%
    transmute(
      Cohort,
      Age,
      Age.group,
      Total,
      Symptom = "Both",
      Type = "Co-occurring",
      Cases = Both
    )
  
) %>%
  
  filter(
    !is.na(Cases),
    !is.na(Total),
    Total > 0
  )

prev_dat$Symptom <- factor(
  prev_dat$Symptom,
  levels = c("HMB", "MP", "Both")
)

prev_dat$Type <- factor(
  prev_dat$Type,
  levels = c("Any", "Isolated", "Co-occurring")
)

ci <- binom::binom.wilson(
  prev_dat$Cases,
  prev_dat$Total
)

prev_dat <- bind_cols(prev_dat, ci)

make_plot <- function(agegroup, title_text, xlab_text = "") {
  
  ggplot(
    prev_dat[prev_dat$Age.group == agegroup, ],
    aes(
      x = Age,
      y = mean,
      colour = Type,
      group = Type
    )
  ) +
    
    geom_line(size = 0.5) +
    
    geom_point(size = 1) +
    
    geom_errorbar(
      aes(
        ymin = lower,
        ymax = upper
      ),
      width = 0.2,
      size = 0.5
    ) +
    
    facet_nested(
      Cohort ~ Symptom,
      scales = "free"
    ) +
    
    scale_y_continuous(
      labels = scales::percent
    ) +
    
    scale_colour_manual(
      values = c(
        "Any" = "cornflowerblue",
        "Isolated" = "skyblue2",
        "Co-occurring" = "darkblue"
      )
    ) +
    
    theme_bw() +
    
    theme(
      strip.text.y = element_text(angle = 0),
      panel.spacing = unit(0.01, "lines"),
      plot.title = element_text(
        size = 10,
        face = "bold"
      ),
      axis.title = element_text(size = 10),
      axis.text = element_text(size = 8),
      strip.text = element_text(size = 8),
      plot.margin = margin(2, 2, 2, 2),
      legend.margin = margin(t = -5),
      legend.box.margin = margin(t = -8),
      legend.title = element_blank(),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(size = 0.2)
    ) +
    
    ylab("Prevalence") +
    xlab(xlab_text) +
    ggtitle(title_text)
}

Ado_plot <- make_plot(
  agegroup = "Ado",
  title_text = "Adolescence"
)

Mid_plot <- make_plot(
  agegroup = "Mid",
  title_text = "Mid reproductive life"
)

Late_plot <- make_plot(
  agegroup = "Peri",
  title_text = "Late reproductive life",
  xlab_text = "Age"
)

combined_plot <- (
  Ado_plot /
    Mid_plot /
    Late_plot
) +
  
  plot_layout(
    guides = "collect",
    heights = c(3, 4, 2)
  ) &
  
  theme(
    legend.position = "bottom"
  )

combined_plot

ggsave("point_prev_final.png", plot = combined_plot, width = 9, height = 13, dpi = 500)


```


#For point prev supplementary table
```{r}
library(dplyr)
library(binom)

export_prev <- dat

# Calculate prevalence and Wilson CIs

calc_ci <- function(cases, total, prefix){

  ci <- binom::binom.wilson(cases, total)

  tibble(
    !!paste0(prefix,"_prev")  := ci$mean * 100,
    !!paste0(prefix,"_lower") := ci$lower * 100,
    !!paste0(prefix,"_upper") := ci$upper * 100
  )

}

export_prev <- bind_cols(

  export_prev,

  calc_ci(export_prev$HMB.any, export_prev$Total, "HMB_any"),

  calc_ci(export_prev$MP.any, export_prev$Total, "MP_any"),

  calc_ci(export_prev$Either, export_prev$Total, "Either"),

  calc_ci(export_prev$HMB.only, export_prev$Total, "HMB_only"),

  calc_ci(export_prev$MP.only, export_prev$Total, "MP_only"),

  calc_ci(export_prev$Both, export_prev$Total, "Both"),

  calc_ci(export_prev$Neither, export_prev$Total, "Neither")

)
export_prev <- export_prev %>%

  mutate(

    across(
      ends_with("_prev"),
      ~round(.x,1)
    ),

    across(
      ends_with("_lower"),
      ~round(.x,1)
    ),

    across(
      ends_with("_upper"),
      ~round(.x,1)
    )

  )

export_prev <- dat %>%

  bind_cols(

    calc_ci(dat$HMB.any,   dat$Total, "HMB_any"),
    calc_ci(dat$MP.any,    dat$Total, "MP_any"),
    calc_ci(dat$Either,    dat$Total, "Either"),
    calc_ci(dat$HMB.only,  dat$Total, "HMB_only"),
    calc_ci(dat$MP.only,   dat$Total, "MP_only"),
    calc_ci(dat$Both,      dat$Total, "Both"),
    calc_ci(dat$Neither,   dat$Total, "Neither")

  ) %>%

  mutate(

    `HMB any` = sprintf(
      "%.1f (%.1f-%.1f)",
      HMB_any_prev,
      HMB_any_lower,
      HMB_any_upper
    ),

    `MP any` = sprintf(
      "%.1f (%.1f-%.1f)",
      MP_any_prev,
      MP_any_lower,
      MP_any_upper
    ),

    `Either` = sprintf(
      "%.1f (%.1f-%.1f)",
      Either_prev,
      Either_lower,
      Either_upper
    ),

    `HMB only` = sprintf(
      "%.1f (%.1f-%.1f)",
      HMB_only_prev,
      HMB_only_lower,
      HMB_only_upper
    ),

    `MP only` = sprintf(
      "%.1f (%.1f-%.1f)",
      MP_only_prev,
      MP_only_lower,
      MP_only_upper
    ),

    `Both` = sprintf(
      "%.1f (%.1f-%.1f)",
      Both_prev,
      Both_lower,
      Both_upper
    ),

    `Neither` = sprintf(
      "%.1f (%.1f-%.1f)",
      Neither_prev,
      Neither_lower,
      Neither_upper
    )

  ) %>%

  select(
    Cohort,
    Age.group,
    Age,
    Total,
    `HMB any`,
    `MP any`,
    Either,
    `HMB only`,
    `MP only`,
    Both,
    Neither
  )
write.csv(
  export_prev,
  "point_prevalence_plot_data.csv",
  row.names = FALSE
)
```


#Descriptive table by SES vars
#Functions
#Descriptive (Table 2)
```{r}
extract_age_range <- function(data, prefix = "hmb_mp_age") {

  age_vars <- grep(prefix, names(data), value = TRUE)
  ages <- as.numeric(gsub("[^0-9]", "", age_vars))

  paste0(min(ages, na.rm = TRUE), "–", max(ages, na.rm = TRUE))

}

create_descriptive_table <- function(data, vars){

  sample_size <- nrow(data)
  age_range <- extract_age_range(data)

  sample_info <- tibble(
    Variable = c("Sample size", "Age range"),
    Category = "",
    Value = c(sample_size, age_range)
  )

  var_tables <- map_df(vars, function(v){

    data %>%
      filter(!is.na(.data[[v]])) %>%
      count(.data[[v]]) %>%
      mutate(
        percent = round(100*n/sum(n),1),
        Variable = v,
        Category = as.character(.data[[v]]),
        Value = paste0(n," (",percent,")")
      ) %>%
      select(Variable, Category, Value)

  })

  bind_rows(sample_info, var_tables)

}
```


```{r}
abcd_vars <- c("edu_parent_harmonised","income_harmonised","ethnicity_harmonised")

bcs_vars <- c("edu_mother_harmonised","edu_father_harmonised","occ_mother_harmonised","occ_father_harmonised","income_harmonised", "ethnicity_harmonised")

raine_vars <- c("edu_mother_harmonised","edu_father_harmonised","income_harmonised", "ethnicity_harmonised")

alspac_g0_vars <- c("ethnicity_harmonised", "occ_mother_harmonised", "occ_father_harmonised")

alspac_g1_vars <- c("edu_mother_harmonised","edu_father_harmonised","occ_mother_harmonised","occ_father_harmonised","ethnicity_harmonised")

abcd_table1 <- create_descriptive_table(abcd_der, abcd_vars)
bcs_table1 <- create_descriptive_table(bcs_der, bcs_vars)
raine_table1 <- create_descriptive_table(raine_der, raine_vars)
alspac_g0_table1 <- create_descriptive_table(alspac_g0_der, alspac_g0_vars)
alspac_g1_table1 <- create_descriptive_table(alspac_g1_der, alspac_g1_vars)

write.csv(abcd_table1,"ABCD_sample_characteristics.csv",row.names=FALSE)
write.csv(bcs_table1,"BCS70_sample_characteristics.csv",row.names=FALSE)
write.csv(raine_table1,"RAINE_sample_characteristics.csv",row.names=FALSE)
write.csv(alspac_g0_table1,"ALSPAC_G0_sample_characteristics.csv",row.names=FALSE)
write.csv(alspac_g1_table1,"ALSPAC_G1_sample_characteristics.csv",row.names=FALSE)

```

#Statified analysis by SES
```{r}
library(dplyr)

abcd_der$hmb_mp_ever <- apply(
  abcd_der[, grep("^hmb_mp_age", names(abcd_der))],
  1,
  function(x) if(all(is.na(x))) NA else max(x, na.rm = TRUE)
)

bcs_der$hmb_mp_ever <- apply(
  bcs_der[, grep("^hmb_mp_age", names(bcs_der))],
  1,
  function(x) if(all(is.na(x))) NA else max(x, na.rm = TRUE)
)

raine_der$hmb_mp_ever <- apply(
  raine_der[, grep("^hmb_mp_age", names(raine_der))],
  1,
  function(x) if(all(is.na(x))) NA else max(x, na.rm = TRUE)
)

alspac_g1_der$hmb_mp_ever <- apply(
  alspac_g1_der[, grep("^hmb_mp_age", names(alspac_g1_der))],
  1,
  function(x) if(all(is.na(x))) NA else max(x, na.rm = TRUE)
)

alspac_g0_der$hmb_mp_ever <- apply(
  alspac_g0_der[, grep("^hmb_mp_age", names(alspac_g0_der))],
  1,
  function(x) if(all(is.na(x))) NA else max(x, na.rm = TRUE)
)

make_symptom_table <- function(data, vars, outcome){

  results <- list()

  for(v in vars){

    df <- data %>%
      filter(
        !is.na(.data[[v]]),
        !is.na(.data[[outcome]])
      )

    tab <- df %>%
      group_by(.data[[v]]) %>%
      summarise(

        N = n(),

        HMB = paste0(
          sum(.data[[outcome]] %in% c(1,3)),
          " (",
          round(
            100 * mean(.data[[outcome]] %in% c(1,3)),
            1
          ),
          ")"
        ),

        MP = paste0(
          sum(.data[[outcome]] %in% c(2,3)),
          " (",
          round(
            100 * mean(.data[[outcome]] %in% c(2,3)),
            1
          ),
          ")"
        ),

        Either = paste0(
          sum(.data[[outcome]] %in% c(1,2,3)),
          " (",
          round(
            100 * mean(.data[[outcome]] %in% c(1,2,3)),
            1
          ),
          ")"
        ),

        HMB_only = paste0(
          sum(.data[[outcome]] == 1),
          " (",
          round(
            100 * mean(.data[[outcome]] == 1),
            1
          ),
          ")"
        ),

        MP_only = paste0(
          sum(.data[[outcome]] == 2),
          " (",
          round(
            100 * mean(.data[[outcome]] == 2),
            1
          ),
          ")"
        ),

        Both = paste0(
          sum(.data[[outcome]] == 3),
          " (",
          round(
            100 * mean(.data[[outcome]] == 3),
            1
          ),
          ")"
        ),

        Neither = paste0(
          sum(.data[[outcome]] == 0),
          " (",
          round(
            100 * mean(.data[[outcome]] == 0),
            1
          ),
          ")"
        ),

        .groups = "drop"
      )

    names(tab)[1] <- "Category"

    tab$Variable <- v

    tab <- tab %>%
      select(
        Variable,
        Category,
        N,
        HMB,
        MP,
        Either,
        HMB_only,
        MP_only,
        Both,
        Neither
      )

    results[[v]] <- tab
  }

  bind_rows(results)
}

abcd_vars <- c(
  "edu_parent_harmonised",
  "income_harmonised",
  "ethnicity_harmonised"
)

bcs_vars <- c(
  "edu_mother_harmonised",
  "edu_father_harmonised",
  "occ_mother_harmonised",
  "occ_father_harmonised",
  "income_harmonised",
  "ethnicity_harmonised"
)

raine_vars <- c(
  "edu_mother_harmonised",
  "edu_father_harmonised",
  "income_harmonised",
  "ethnicity_harmonised"
)

g0_vars <- c(
  "ethnicity_harmonised",
  "occ_mother_harmonised",
  "occ_father_harmonised"
)

g1_vars <- c(
  "edu_mother_harmonised",
  "edu_father_harmonised",
  "occ_mother_harmonised",
  "occ_father_harmonised",
  "ethnicity_harmonised"
)

abcd_sep_table <- make_symptom_table(
  abcd_der,
  abcd_vars,
  "hmb_mp_ever"
)

bcs_sep_table <- make_symptom_table(
  bcs_der,
  bcs_vars,
  "hmb_mp_ever"
)

raine_sep_table <- make_symptom_table(
  raine_der,
  raine_vars,
  "hmb_mp_ever"
)

alspac_g0_sep_table <- make_symptom_table(
  alspac_g0_der,
  g0_vars,
  "hmb_mp_ever"
)

alspac_g1_sep_table <- make_symptom_table(
  alspac_g1_der,
  g1_vars,
  "hmb_mp_ever"
)

write.csv(
  abcd_sep_table,
  "ABCD_SEP_symptoms.csv",
  row.names = FALSE
)

write.csv(
  bcs_sep_table,
  "BCS70_SEP_symptoms.csv",
  row.names = FALSE
)

write.csv(
  raine_sep_table,
  "RAINE_SEP_symptoms.csv",
  row.names = FALSE
)

write.csv(
  alspac_g0_sep_table,
  "ALSPAC_G0_SEP_symptoms.csv",
  row.names = FALSE
)

write.csv(
  alspac_g1_sep_table,
  "ALSPAC_G1_SEP_symptoms.csv",
  row.names = FALSE
)
```

#SEP table for plotting
```{r}
library(dplyr)
library(tidyr)
library(stringr)

make_export_table <- function(tab, cohort){

  tab %>%

    mutate(

      Cohort = cohort,

      Demographic = recode(
        Variable,
        edu_parent_harmonised = "Parental education",
        edu_mother_harmonised = "Mother's education",
        edu_father_harmonised = "Father's education",
        occ_mother_harmonised = "Mother's occupation",
        occ_father_harmonised = "Father's occupation",
        income_harmonised = "Household income"
      )

    ) %>%

    # Remove ethnicity
    filter(Variable != "ethnicity_harmonised") %>%

    mutate(

      Level = case_when(
        Category %in% c(1, "1", "low", "Low") ~ "Low",
        Category %in% c(2, "2", "medium", "Medium") ~ "Medium",
        Category %in% c(3, "3", "high", "High") ~ "High",
        TRUE ~ as.character(Category)
      )

    ) %>%

    pivot_longer(
      cols = c(HMB, MP, Either),
      names_to = "Symptom",
      values_to = "Value"
    ) %>%

    mutate(
      Cases = as.numeric(stringr::str_extract(Value, "^[0-9]+")),
      Total = N
    ) %>%

    select(
      Cohort,
      Symptom,
      Demographic,
      Level,
      Total,
      Cases
    )

}

abcd_export <- make_export_table(abcd_sep_table, "ABCD")

g1_export <- make_export_table(alspac_g1_sep_table, "ALSPAC G1")

raine_export <- make_export_table(raine_sep_table, "Raine")

bcs_export <- make_export_table(bcs_sep_table, "BCS70")

g0_export <- make_export_table(alspac_g0_sep_table, "ALSPAC G0")
g0_manual <- tribble(
~Cohort, ~Symptom, ~Demographic, ~Level, ~Total, ~Cases,

# Mother's education
"ALSPAC G0","HMB","Mother's education","Low",4963,4460,
"ALSPAC G0","HMB","Mother's education","Medium",2381,2170,
"ALSPAC G0","HMB","Mother's education","High",365,327,

"ALSPAC G0","MP","Mother's education","Low",4963,3065,
"ALSPAC G0","MP","Mother's education","Medium",2381,1440,
"ALSPAC G0","MP","Mother's education","High",365,207,

"ALSPAC G0","Either","Mother's education","Low",4963,4535,
"ALSPAC G0","Either","Mother's education","Medium",2381,2196,
"ALSPAC G0","Either","Mother's education","High",365,337,

# Father's education
"ALSPAC G0","HMB","Father's education","Low",4516,4064,
"ALSPAC G0","HMB","Father's education","Medium",1898,1731,
"ALSPAC G0","HMB","Father's education","High",899,817,

"ALSPAC G0","MP","Father's education","Low",4516,2764,
"ALSPAC G0","MP","Father's education","Medium",1898,1188,
"ALSPAC G0","MP","Father's education","High",899,519,

"ALSPAC G0","Either","Father's education","Low",4516,4129,
"ALSPAC G0","Either","Father's education","Medium",1898,1751,
"ALSPAC G0","Either","Father's education","High",899,832
)

g0_export <- bind_rows(
  g0_export,
  g0_manual
)

moba_export <- tribble(
~Cohort, ~Symptom, ~Demographic, ~Level, ~Total, ~Cases,

"MoBa","HMB","Mother's education","Low",286,78,
"MoBa","HMB","Mother's education","Medium",1257,317,
"MoBa","HMB","Mother's education","High",3336,813,

"MoBa","HMB","Father's education","Low",436,111,
"MoBa","HMB","Father's education","Medium",1723,414,
"MoBa","HMB","Father's education","High",2491,624,

"MoBa","HMB","Parental education","Low",NA,NA,
"MoBa","HMB","Parental education","Medium",NA,NA,
"MoBa","HMB","Parental education","High",NA,NA,

"MoBa","HMB","Mother's occupation","Low",NA,NA,
"MoBa","HMB","Mother's occupation","Medium",NA,NA,
"MoBa","HMB","Mother's occupation","High",NA,NA,

"MoBa","HMB","Father's occupation","Low",NA,NA,
"MoBa","HMB","Father's occupation","Medium",NA,NA,
"MoBa","HMB","Father's occupation","High",NA,NA,

"MoBa","HMB","Household income","Low",1935,488,
"MoBa","HMB","Household income","Medium",1159,271,
"MoBa","HMB","Household income","High",1469,368,

"MoBa","MP","Mother's education","Low",286,219,
"MoBa","MP","Mother's education","Medium",1257,934,
"MoBa","MP","Mother's education","High",3336,2453,

"MoBa","MP","Father's education","Low",436,334,
"MoBa","MP","Father's education","Medium",1723,1279,
"MoBa","MP","Father's education","High",2491,1829,

"MoBa","MP","Parental education","Low",NA,NA,
"MoBa","MP","Parental education","Medium",NA,NA,
"MoBa","MP","Parental education","High",NA,NA,

"MoBa","MP","Mother's occupation","Low",NA,NA,
"MoBa","MP","Mother's occupation","Medium",NA,NA,
"MoBa","MP","Mother's occupation","High",NA,NA,

"MoBa","MP","Father's occupation","Low",NA,NA,
"MoBa","MP","Father's occupation","Medium",NA,NA,
"MoBa","MP","Father's occupation","High",NA,NA,

"MoBa","MP","Household income","Low",1935,1460,
"MoBa","MP","Household income","Medium",1159,851,
"MoBa","MP","Household income","High",1469,1068,

"MoBa","Either","Mother's education","Low",286,231,
"MoBa","Either","Mother's education","Medium",1257,971,
"MoBa","Either","Mother's education","High",3336,2570,

"MoBa","Either","Father's education","Low",436,345,
"MoBa","Either","Father's education","Medium",1723,1332,
"MoBa","Either","Father's education","High",2491,1921,

"MoBa","Either","Parental education","Low",NA,NA,
"MoBa","Either","Parental education","Medium",NA,NA,
"MoBa","Either","Parental education","High",NA,NA,

"MoBa","Either","Mother's occupation","Low",NA,NA,
"MoBa","Either","Mother's occupation","Medium",NA,NA,
"MoBa","Either","Mother's occupation","High",NA,NA,

"MoBa","Either","Father's occupation","Low",NA,NA,
"MoBa","Either","Father's occupation","Medium",NA,NA,
"MoBa","Either","Father's occupation","High",NA,NA,

"MoBa","Either","Household income","Low",1935,1521,
"MoBa","Either","Household income","Medium",1159,885,
"MoBa","Either","Household income","High",1469,1122
)

sep_export <- bind_rows(
  abcd_export,
  g1_export,
  raine_export,
  moba_export,
  bcs_export,
  g0_export
)

sep_export <- sep_export %>%

  mutate(

    Cohort = factor(
      Cohort,
      levels = c(
        "ABCD",
        "ALSPAC G1",
        "Raine",
        "MoBa",
        "BCS70",
        "ALSPAC G0"
      )
    ),

    Symptom = factor(
      Symptom,
      levels = c("HMB", "MP", "Either")
    ),

    Demographic = factor(
      Demographic,
      levels = c(
        "Mother's education",
        "Father's education",
        "Parental education",
        "Mother's occupation",
        "Father's occupation",
        "Household income"
      )
    ),

    Level = factor(
      Level,
      levels = c("Low", "Medium", "High")
    )

  ) %>%

  arrange(
    Cohort,
    Symptom,
    Demographic,
    Level
  )

write.csv(sep_export,"SEP_prevalence_counts.csv",row.names = FALSE)
```


#Prevalence ratios (Figure S4)
```{r}
dat <- sep_export
library(writexl)

# Calculate prevalence
dat2 <- dat %>%
  mutate(prev = Cases / Total)

# Get prevalence in High group (reference)
pr_dat <- dat2 %>%
  select(Cohort, Symptom, Demographic, Level, prev, Cases, Total) %>%
  left_join(
    dat2 %>%
      filter(Level == "High") %>%
      select(
        Cohort, Symptom, Demographic,
        prev_ref = prev,
        Cases_ref = Cases,
        Total_ref = Total
      ),
    by = c("Cohort", "Symptom", "Demographic")
  ) %>%
  mutate(
    PR = prev / prev_ref,
    
    # Wald CI on log(PR)
    se_logPR = sqrt(
      (1/Cases) - (1/Total) +
        (1/Cases_ref) - (1/Total_ref)
    ),
    
    lower = exp(log(PR) - 1.96 * se_logPR),
    upper = exp(log(PR) + 1.96 * se_logPR)
  )

pr_dat$lower[pr_dat$Level=="High"]<-1
pr_dat$upper[pr_dat$Level=="High"]<-1

pr_dat$Level <- factor(pr_dat$Level,ordered=T,levels=c("Low","Medium","High"))

pr_table <- pr_dat %>%
  mutate(
    PR = round(PR, 2),
    lower = round(lower, 2),
    upper = round(upper, 2),
    `PR (95% CI)` = sprintf("%.2f (%.2f–%.2f)", PR, lower, upper)
  ) %>%
  select(
    Cohort,
    Symptom,
    Demographic,
    Level,
    Cases,
    Total,
    prev,
    PR,
    lower,
    upper,
    `PR (95% CI)`
  )

write_xlsx(
  pr_table,
  "Prevalence_ratios_with_CIs.xlsx"
)

ggplot(
  na.omit(pr_dat),
  aes(
    y = Level,
    x = PR,
    xmin = lower,
    xmax = upper,
    colour=Level
  )
) +
  geom_vline(xintercept = 1, linetype = 2,colour="grey60") +
  geom_pointrange(
    size=0.2,
    position = position_dodge(width = 0.4)
  ) +
  facet_nested(Cohort+Demographic ~ Symptom,scales="free",space="free_y", switch="y", strip = strip_nested(size = "variable")) +
  theme_light() +
  scale_colour_manual(values=c('Low'="#cc4c02",'Medium'="#fe9929",'High'="#fed98e"))+
  labs(
    y = NULL,
    x = "Prevalence ratio (vs High)"
  )+
  theme(strip.text.y.left = element_text(angle = 0),
        panel.grid.major=element_blank(), panel.grid.minor=element_blank(),
        axis.text.y = element_blank(),axis.ticks.y=element_blank(),
        legend.position = "bottom",panel.spacing = unit(0.1, "lines"))

ggsave("prevalence_ratio.png", width = 10, height = 12, dpi = 500)

```

##Prevelance by SES bar plot (Figure S3)
```{r}
#Calculate prevalence
dat <- dat %>%
  mutate(
    Prevalence = (Cases / Total) * 100
  )

#Order symptoms
dat <- dat %>%
  mutate(Symptom = factor(Symptom, levels = c("HMB", "MP", "Either"))
  )

#Order demo factors
dat$Demographic <- factor(
  dat$Demographic,
  levels = c(
    "Mother's education",
    "Father's education",
    "Parental education",
    "Household income",
    "Mother's occupation",
    "Father's occupation")
)

#Order demo levels
dat <- dat %>%
  mutate(Level = factor(Level, levels = c("Low", "Medium", "High"))
  )

#Plot 
ggplot(
  dat,
  aes(x = Symptom, y = Prevalence, fill = Level)
) +
  scale_y_continuous(
    limits = c(0, 110), #extra space on y axis for vertical prevalence text
    breaks = seq(0, 100, by = 20)
  ) +
  geom_col(
    position = position_dodge(width = 0.8), width = 0.7
  ) +
  geom_text(
    aes(label = sprintf("%.1f", Prevalence)),
    position = position_dodge(width = 0.8),
    vjust = 0.4, hjust = -0.05, angle = 90, size = 2.7
  ) +
  facet_grid(
    Cohort ~ Demographic, scales = "free_x", space = "free_x"
  ) +
  labs(
    x = NULL, y = "Prevalence (%)"
  ) +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "grey95"),
    strip.text.x = element_text(face = "bold", size=8),
    strip.text.y = element_text(face = "bold", size=9),
    axis.text.x = element_text(size = 9)
  ) +
  scale_fill_manual(
    values = c(
      "Low" = "#cc4c02",
      "Medium" = "#fe9929",
      "High" = "#fed98e"
    )
  )

ggsave("ses_bar_plot.png", width = 10, height = 12, dpi = 500)

```

#Ethnicity (Figure S5)
```{r}
ethnicity_plot <- tribble(
  ~Cohort, ~Ethnicity, ~HMB, ~MP, ~Either,

  # ABCD
  "ABCD", "White", 32.9, 59.4, 66.3,
  "ABCD", "Black", 32.4, 75.1, 79.5,
  "ABCD", "Asian", 27.7, 47.9, 57.4,
  "ABCD", "Other", 35.1, 65.6, 72.5,

  # ALSPAC G1
  "ALSPAC G1", "White", 58.0, 72.4, 81.6,
  "ALSPAC G1", "Non-white", 66.7, 73.1, 83.3,

  # Raine
  "Raine", "White", 40.3, 64.8, 75.5,
  "Raine", "Non-white", 38.4, 52.3, 69.8,

    # BCS70
  "BCS70", "White", 58.5, 48.4, 66.1,
  "BCS70", "Non-white", 56.9, 53.8, 67.7,

  # ALSPAC G0
  "ALSPAC G0", "White", 89.9, 61.1, 92.9,
  "ALSPAC G0", "Non-white", 93.4, 60.2, 95.3
)
library(tidyr)

ethnicity_long <- ethnicity_plot %>%
  pivot_longer(
    cols = c(HMB, MP, Either),
    names_to = "Symptom",
    values_to = "Prevalence"
  )

library(ggplot2)
library(dplyr)

ethnicity_long$Ethnicity <- factor(
  ethnicity_long$Ethnicity,
  levels = c("White","Black","Asian","Other","Non-white")
)

ethnicity_long$Cohort <- factor(
  ethnicity_long$Cohort,
  levels = c(
    "ABCD",
    "ALSPAC G1",
    "Raine",
    "BCS70",
    "ALSPAC G0"
  )
)

ethnicity_plot <- ggplot(
  ethnicity_long,
  aes(
    x = Symptom,
    y = Prevalence,
    fill = Ethnicity
  )
) +

  geom_col(
    position = position_dodge(width = 0.8),
    width = 0.7
  ) +

  geom_text(
    aes(
      label = round(Prevalence, 1)
    ),
    position = position_dodge(width = 0.8),
    vjust = -0.25,
    angle = 0,
    size = 3
  ) +

  facet_wrap(
    ~ Cohort,
    nrow = 1
  ) +

  scale_y_continuous(
    limits = c(0, 110),
    expand = expansion(mult = c(0, 0.02))
  ) +

  labs(
    x = NULL,
    y = "Prevalence (%)",
    fill = "Ethnicity"
  ) +

  theme_bw(base_size = 12) +

  theme(
    legend.position = "bottom",

    strip.background = element_rect(
      fill = "grey90",
      colour = "black"
    ),

    strip.text = element_text(
      face = "bold"
    ),

    panel.grid.major.x = element_blank(),

    axis.text.x = element_text(
      angle = 0,
      hjust = 0.5
    )
  )

ethnicity_plot

ggsave("ethnicity_prevalence_plot.png", ethnicity_plot, width = 14, height = 7, dpi = 300)

library(dplyr)
library(writexl)
library(tibble)

eth <- tribble(
  ~Cohort, ~Ethnicity, ~Symptom, ~Cases, ~Total,

  # ABCD
"ABCD","White","HMB",878,2665,
"ABCD","Black","HMB",250,772,
"ABCD","Asian","HMB",26,94,
"ABCD","Others","HMB",545,1554,

"ABCD","White","MP",1582,2665,
"ABCD","Black","MP",580,772,
"ABCD","Asian","MP",45,94,
"ABCD","Others","MP",1020,1554,

"ABCD","White","Either",1768,2665,
"ABCD","Black","Either",614,772,
"ABCD","Asian","Either",54,94,
"ABCD","Others","Either",1126,1554,

  # ALSPAC G1
  "ALSPAC G1","White","HMB",2095,3610,
  "ALSPAC G1","Non-white","HMB",104,156,

  "ALSPAC G1","White","MP",2614,3610,
  "ALSPAC G1","Non-white","MP",113,156,

  "ALSPAC G1","White","Either",2945,3610,
  "ALSPAC G1","Non-white","Either",130,156,

  # Raine
  "Raine","White","HMB",220,546,
  "Raine","Non-white","HMB",33,86,

  "Raine","White","MP",354,546,
  "Raine","Non-white","MP",45,86,

  "Raine","White","Either",412,546,
  "Raine","Non-white","Either",60,86,

  # BCS70
"BCS70","White","HMB",1762,3014,
"BCS70","Non-white","HMB",37,65,

"BCS70","White","MP",1459,3014,
"BCS70","Non-white","MP",35,65,

"BCS70","White","Either",1993,3014,
"BCS70","Non-white","Either",44,65,

# ALSPAC G0
"ALSPAC G0","White","HMB",8833,9832,
"ALSPAC G0","Non-white","HMB",197,211,

"ALSPAC G0","White","MP",6003,9832,
"ALSPAC G0","Non-white","MP",127,211,

"ALSPAC G0","White","Either",9129,9832,
"ALSPAC G0","Non-white","Either",201,211
)

eth <- eth %>%
  mutate(prev = Cases / Total)

eth_pr <- eth %>%
  left_join(
    eth %>%
      filter(Ethnicity == "White") %>%
      select(
        Cohort,
        Symptom,
        prev_ref = prev,
        Cases_ref = Cases,
        Total_ref = Total
      ),
    by = c("Cohort","Symptom")
  ) %>%
  mutate(
    PR = prev / prev_ref,

    se_logPR = sqrt(
      (1/Cases) - (1/Total) +
      (1/Cases_ref) - (1/Total_ref)
    ),

    lower = exp(log(PR) - 1.96 * se_logPR),
    upper = exp(log(PR) + 1.96 * se_logPR)
  )

## Reference group
eth_pr <- eth_pr %>%
  mutate(
    PR = ifelse(Ethnicity == "White", 1, PR),
    lower = ifelse(Ethnicity == "White", 1, lower),
    upper = ifelse(Ethnicity == "White", 1, upper)
  )

eth_table <- eth_pr %>%
  mutate(
    Prevalence = round(prev*100,1),
    PR = round(PR,2),
    lower = round(lower,2),
    upper = round(upper,2),
    `PR (95% CI)` = sprintf("%.2f (%.2f–%.2f)", PR, lower, upper)
  ) %>%
  select(
    Cohort,
    Symptom,
    Ethnicity,
    Cases,
    Total,
    Prevalence,
    `PR (95% CI)`
  )

write_xlsx(
  eth_table,
  "Ethnicity_prevalence_ratios.xlsx"
)

```

#Incidence analysis
# Incidence by gynaecological age (Figure 5)

```{r}
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)

# Import incidence tables exported from Stata
abcd_inc <- read_excel("C:/Users/sk1014/OneDrive - University of Exeter/Sharp, Gemma's files - Shared/Analysis plans/Analysis/Aim 1/abcd_incidence.xlsx") %>%
  mutate(cohort = "ABCD")

g1_inc <- read_excel("C:/Users/sk1014/OneDrive - University of Exeter/Sharp, Gemma's files - Shared/Analysis plans/Analysis/Aim 1/alspac_g1_incidence.xlsx") %>%
  mutate(cohort = "ALSPAC G1")

abcd_inc <- abcd_inc %>%
  mutate(
    cohort = "ABCD")

g1_inc <- g1_inc %>%
  mutate(
    cohort = "ALSPAC G1")

combined_inc <- bind_rows(abcd_inc, g1_inc)

# Combine cohorts
combined_inc <- bind_rows(abcd_inc,g1_inc)

# Keep post-menarche only
combined_inc <- combined_inc %>%
  filter(ts_menarche >= 0)

library(binom)

# Wilson CIs
hmb_ci <- binom.wilson(
  combined_inc$incident_hmb,
  combined_inc$risk_hmb
)

mp_ci <- binom.wilson(
  combined_inc$incident_mp,
  combined_inc$risk_mp
)

combined_inc <- combined_inc %>%
  mutate(
    hmb_lower = hmb_ci$lower,
    hmb_upper = hmb_ci$upper,
    mp_lower  = mp_ci$lower,
    mp_upper  = mp_ci$upper
  )

combined_inc_long <- bind_rows(

  combined_inc %>%
    transmute(
      cohort,
      ts_half,
      type = "HMB ever",
      incidence = hmb_incidence,
      lower = hmb_lower,
      upper = hmb_upper
    ),

  combined_inc %>%
    transmute(
      cohort,
      ts_half,
      type = "MP ever",
      incidence = mp_incidence,
      lower = mp_lower,
      upper = mp_upper
    )
)

combined_inc_ts <- combined_inc

ggplot(
  combined_inc_long,
  aes(
    x = ts_half,
    y = incidence*100,
    colour = type
  )
) +

  geom_line(linewidth = 1) +
  geom_point(size = 1.8) +
  geom_errorbar(
  aes(
    ymin = lower * 100,
    ymax = upper * 100
  ),
  width = 0.15,
  linewidth = 0.4
) +

  geom_vline(
    xintercept = 0,
    linetype = "dashed"
  ) +

  facet_wrap(
    ~cohort,
    scales = "free_x"
  ) +

  scale_color_manual(
    values = c(
      "HMB ever" = "black",
      "MP ever" = "grey50"
    )
  ) +

  labs(
    x = "Time since menarche (years)",
    y = "Incidence (%)",
    color = "Symptom"
  ) +

  theme_minimal(base_size = 14) +

theme(
  legend.position = "bottom"
)

ggsave(
  "incidence_menarche_plot.png",
  width = 12,
  height = 5,
  dpi = 300
)
```


#Incidence by chronological age (Figure 4)

```{r}

# Import incidence tables exported from Stata
abcd_inc <- read_excel("C:/Users/sk1014/OneDrive - University of Exeter/Sharp, Gemma's files - Shared/Analysis plans/Analysis/Aim 1/abcd_incidence_age.xlsx") %>%  
  rename(age = age_year)

g1_inc <- read_excel("C:/Users/sk1014/OneDrive - University of Exeter/Sharp, Gemma's files - Shared/Analysis plans/Analysis/Aim 1/alspac_g1_incidence_age.xlsx") %>% 
  rename(age = age_year)
  

abcd_inc <- abcd_inc %>%
  mutate(
    cohort = "ABCD")

g1_inc <- g1_inc %>%
  mutate(
    cohort = "ALSPAC G1")

combined_inc <- bind_rows(abcd_inc, g1_inc)
combined_inc <- combined_inc %>%
  mutate(
    hmb_incidence = ifelse(risk_hmb < 5, NA, hmb_incidence),
    mp_incidence  = ifelse(risk_mp  < 5, NA, mp_incidence)
  )
library(binom)

# HMB confidence intervals
hmb_ci <- binom.wilson(
  combined_inc$incident_hmb,
  combined_inc$risk_hmb
)

# MP confidence intervals
mp_ci <- binom.wilson(
  combined_inc$incident_mp,
  combined_inc$risk_mp
)

combined_inc <- combined_inc %>%
  mutate(
    hmb_lower = hmb_ci$lower,
    hmb_upper = hmb_ci$upper,
    mp_lower  = mp_ci$lower,
    mp_upper  = mp_ci$upper
  )

combined_inc_long_age <- bind_rows(

  combined_inc %>%
    transmute(
      cohort,
      age,
      type = "HMB ever",
      incidence = hmb_incidence,
      lower = hmb_lower,
      upper = hmb_upper
    ),

  combined_inc %>%
    transmute(
      cohort,
      age,
      type = "MP ever",
      incidence = mp_incidence,
      lower = mp_lower,
      upper = mp_upper
    )

)

combined_inc_age <- combined_inc

combined_inc_long_age$type <- factor(
  combined_inc_long_age$type,
  levels = c("HMB ever","MP ever")
)
combined_inc_long_age <- combined_inc_long_age %>%
  mutate(
    age_plot = case_when(
      cohort == "ALSPAC G1" & age == 10 ~ "9–10",
      cohort == "ALSPAC G1" & age == 11 ~ "11–12",
      TRUE ~ as.character(age)
    )
  )

# Keep ages in the correct order
combined_inc_long_age$age_plot <- factor(
  combined_inc_long_age$age_plot,
  levels = c(
    "9–10", "11–12",
    "11","12","13","14","15","16","17","18"
  )
)

age_plot <- ggplot(
  combined_inc_long_age,
  aes(
    x = age_plot,
    y = incidence * 100,
    colour = type,
    group = type
  )
) +

  geom_line(linewidth = 1) +

  geom_point(size = 2) +
  geom_errorbar(
  aes(
    ymin = lower * 100,
    ymax = upper * 100
  ),
  width = 0.15,
  linewidth = 0.4
) +

  facet_wrap(
    ~cohort,
    scales = "free_x"
  ) +

  scale_color_manual(
    values = c(
      "HMB ever" = "black",
      "MP ever"  = "grey50"
    )
  )  +

  labs(
    x = "Age (years)",
    y = "Incidence (%)",
    colour = "Symptom"
  ) +

  theme_minimal(base_size = 14) +

  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )

age_plot

ggsave(
  "incidence_age_plot.png",
  age_plot,
  width = 12,
  height = 5,
  dpi = 300
)
```

#Incidence table
#Supplementary table - Incidence by time since menarche
```{r}

incidence_table_ts <- combined_inc_ts %>%

  arrange(cohort, ts_half) %>%

  mutate(

    Time_since_menarche = case_when(
      cohort == "ABCD" & ts_half == 8 ~ "8+",
      cohort == "ALSPAC G1" & ts_half == 6 ~ "6+",
      TRUE ~ sprintf("%.1f", ts_half)
    ),

    HMB_summary = sprintf(
      "%d/%d (%.1f%%; %.1f-%.1f)",
      incident_hmb,
      risk_hmb,
      hmb_incidence_pct,
      hmb_lower * 100,
      hmb_upper * 100
    ),

    MP_summary = sprintf(
      "%d/%d (%.1f%%; %.1f-%.1f)",
      incident_mp,
      risk_mp,
      mp_incidence_pct,
      mp_lower * 100,
      mp_upper * 100
    )

  ) %>%

  select(
    cohort,
    Time_since_menarche,
    HMB_summary,
    MP_summary
  )

incidence_table_ts

write.csv(
  incidence_table_ts,
  "incidence_supplementary_table_gynaecological_age.csv",
  row.names = FALSE
)
```

#Supplementary table - Incidence by chronological age
```{r, echo=TRUE, results='asis', fig.show='hold'}

incidence_table_age <- combined_inc_age %>%

  arrange(cohort, age) %>%

  mutate(

    Age_group = case_when(

      cohort == "ALSPAC G1" & age == 10 ~ "9-10",

      cohort == "ALSPAC G1" & age == 11 ~ "11-12",

      TRUE ~ as.character(age)

    ),

    HMB_summary = sprintf(
      "%d/%d (%.1f%%; %.1f-%.1f)",
      incident_hmb,
      risk_hmb,
      hmb_incidence_pct,
      hmb_lower * 100,
      hmb_upper * 100
    ),

    MP_summary = sprintf(
      "%d/%d (%.1f%%; %.1f-%.1f)",
      incident_mp,
      risk_mp,
      mp_incidence_pct,
      mp_lower * 100,
      mp_upper * 100
    )

  ) %>%

  select(
    cohort,
    Age_group,
    HMB_summary,
    MP_summary
  )

incidence_table_age

write.csv(
  incidence_table_age,
  "incidence_supplementary_table_chronological_age.csv",
  row.names = FALSE
)

```
